PanGraph: scalable bacterial pan-genome graph construction

### 2022-09-05 main.tex
Update on Overleaf.

\paragraph{Pairwise alignment.}
Full genome alignment between two closely related isolates is a well-studied problem with many high-quality tools available \cite{li2018minimap2,marccais2018mummer4}.
We chose to use \emph{minimap2} as the core pairwise genome aligner for its proven speed, sensitivity, and easy-to-use, exposed library API \cite{li2018minimap2}.
However, we note that \emph{PanGraph} is written to be modular; in the future, additional alignment kernels can be added with ease.
The \emph{minimap2} alignment kernel is included within a custom Julia wrapper, available at \url{github.com/nnoll/minimap2_jll.jl}.
The kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.

@@

\section{Validation and performance}
In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $1e-2$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{paper-accuracy-20.png}
    \caption{{\bf Accuracy against synthetic data}. Cumulative density functions for breakpoint accuracy comparing simulated and reconstructed pangenome graphs for various SNP rates, $\mu$.}
    \label{fig:toy-accuracy}
\end{figure}

We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, shown in Fig. \ref{fig:toy-performance} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{benchmark.png}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation respectively.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
The mean and standard deviation of results obtained from 25 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.

@@

\begin{figure}[htb]
    \includegraphics[width=.5\textwidth]{panx-compare.png}
    \caption{{\bf Pancontigs cluster towards gene boundaries}.
    The distribution of the nearest gene boundary for each pancontig breakpoint, shown for five species, with samples obtained from RefSeq \cite{o2016reference}.
    (Left) Empirical distribution obtained over all genomes.
    (Right) Null distribution constructed by randomly sampling positions on each genome, constrained to conserve the number of breakpoints.
    }
    \label{fig:panx-accuracy}
\end{figure}

@@

>>>>

\paragraph{Pairwise alignment.}
Full genome alignment between two closely related isolates is a well-studied problem with many high-quality tools available \cite{li2018minimap2,marccais2018mummer4}.
We chose to use \emph{minimap2} as the core pairwise genome aligner for its proven speed, sensitivity, and easy-to-use exposed library API \cite{li2018minimap2}.
This alignment kernel is included within a custom Julia wrapper, available at \url{github.com/nnoll/minimap2_jll.jl}.
However, we note that \emph{PanGraph} is written to be modular, and additional alignment kernels can be added with ease.
In particular we decided to include the option to use \emph{mmseqs2} \cite{steinegger2017mmseqs2} as an alternative alignment kernel, because of its sensitivity on highly diverged sequences at the cost of higher computational time.
In general, the kernel interface expects two lists of \emph{pancontigs} to align as input and will output a list of putative overlapping alignments between both input lists.

@@

\section{Validation and performance}

\subsection{Validation on synthetic data}

In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/median_misplacement_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}}
    \label{fig:toy-accuracy}
\end{figure}

We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, shown in Fig. \ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{benchmark.pdf}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation respectively.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.

\subsection{Validation on real data}

@@

\begin{figure*}[htb]
    \includegraphics[width=.8\textwidth]{figs/incr_size.pdf}
    \caption{{\bf Pangenome graph properties vs. dataset size}}
    \label{fig:panx-size}
\end{figure*}

\begin{figure*}[htb]
    \includegraphics[width=.7\textwidth]{figs/panx_benchmark.pdf}
    \caption{{\bf Benchmark on real data}.}
    \label{fig:panx-benchmark}
\end{figure*}

@@

\begin{figure*}[htb]
    \includegraphics[width=.8\textwidth]{figs/proj_single_escherichia_coli.pdf}
    \caption{{\bf Test of graph marginalization}.}
    \label{fig:marginalization}
\end{figure*}
